Galaxy image statistical generator: how to be confident?

### 2024-09-07 paper/main.tex
flows

Independently of the VAE context, a flow $T$ is defined as diffeomorphism (aka bijector) between the data space and a latent space such that
\begin{align}
\bm{x} = T(\bm{z}) & \quad \Leftrightarrow \quad \bm{z} = T^{-1}(\bm{x})
\end{align}
So, if $\bm{z}$ distribution is $\pi(\bm{z})$ (e.g. $\mathcal{N}(\bm{z},\bm{0},\bm{1})$) then the following relation apply
\begin{equation}
p(\bm{x}) =  \pi(\bm{z})|\det J_T(\bm{z})|^{-1} = \pi(T^{-1}(\bm{x}))|\det J_{T^{-1}}(\bm{x})|
\end{equation}
Contrary to VAE and GAN the latent and data space dimensionalities are the same. The flow $T$ is in general a composition of several individual flows $\{T_i\}_{i<n}$ in such a way that
\begin{align}
T=T_1\circ T_2\circ \dots \circ T_n & \quad \Leftrightarrow \quad 
T^{-1}=T_n^{-1}\circ T_{n-1}^{-1}\circ \dots \circ T_1^{-1}
\end{align}
If one notes $\bm{z}_i = T_i(\bm{z}_{i-1})$ for all $i<n$  (using $\bm{z}_0=\bm{z}$, $\bm{z}_n=\bm{x}$) then
\begin{equation}
\log |\mathrm{det} J_T| = \sum_{i=0}^{n-1} \log |\det J_{T_i}(\bm{z}_{i-1})|
\end{equation}
So, if the flows are well chosen, one can sample $\bm{x}$ with potentially complex (multi-modes) density distribution from a simple spherical multivariate Gaussian distribution. Figure \ref{fig-normflow} represents a schematic views of the \textit{forward} or \textit{generative} direction and the \textit{backward} or \textit{training} direction. 
%
\begin{figure}
\resizebox{\columnwidth}{!}{%
%\begin{center}
\begin{tikzpicture}[
node distance = 1cm, 
distrib/.style={rectangle},
]

@@

The computation of the $\{J_{T_i}\}_{i<n}$ determinants may be challenging except if there are triangular matrices which is the case for some generic flows as . \cite{Papamakarios2021} propose a detailed review on the different architectures  of normalizing flows. 
%
\subsection{Differences between the models}
%
%
\section{Experiment}

>>>>

Independently of the VAE context, a flow $T$ is defined as diffeomorphism (aka bijector) between the data space and a latent space such that
\begin{align}
\bm{x} = T(\bm{z}) & \quad \Leftrightarrow \quad \bm{z} = T^{-1}(\bm{x})
\end{align}
So, if $\bm{z}$ distribution is $\pi(\bm{z})$ (e.g. $\mathcal{N}(\bm{z},\bm{0},\bm{1})$) then the following relation apply
\begin{equation}
p(\bm{x}) =  \pi(\bm{z})|\det J_T(\bm{z})|^{-1} = \pi(T^{-1}(\bm{x}))|\det J_{T^{-1}}(\bm{x})|
\end{equation}
Contrary to VAE and GAN the latent and data space dimensionalities are the same. The flow $T$ is in general a composition of several individual flows $\{T_i\}_{i<n}$ in such a way that
\begin{align}
T=T_1\circ T_2\circ \dots \circ T_n & \quad \Leftrightarrow \quad 
T^{-1}=T_n^{-1}\circ T_{n-1}^{-1}\circ \dots \circ T_1^{-1}
\end{align}
Figure \ref{fig-normflow} represents a schematic views of the \textit{forward} or \textit{generative} direction and the \textit{backward} or \textit{training} direction. 
If one notes $\bm{z}_i = T_i(\bm{z}_{i-1})$ for all $i<n$  (using $\bm{z}_0=\bm{z}$, $\bm{z}_n=\bm{x}$) then
\begin{equation}
\log |\mathrm{det} J_T| = \sum_{i=0}^{n-1} \log |\det J_{T_i}(\bm{z}_{i-1})|
\label{eq-flow-jacob}
\end{equation}
Theorem exists showing that under moderate assumptions flow-based models are able to represent any density distributions \citep{Bogachev2005,huang2019solving}. So, if the flows are well chosen, one can sample $\bm{x}$ with potentially complex (multi-modes) density distribution from a simple spherical multivariate Gaussian distribution.

%
\begin{figure}
\resizebox{\columnwidth}{!}{%
%\begin{center}
\begin{tikzpicture}[
node distance = 1cm, 
distrib/.style={rectangle},
]

@@

Let the flow be parametrized by a vector $\bm{\theta}$ leading to a parametrized flow-model $p_{\bm{\theta}}(\bm{x})$. To obtain an optimized flow, we consider the $\DKL$ divergence between  $p_{\bm{\theta}}(\bm{x})$ and the true $p(\bm{x})$ leads to (n.b we consider $\pi(\bm{z})=\mathcal{N}(\bm{z}, \bm{0}, \bm{1})$ without any extra unknown parameters)
\begin{align}
\mathcal{L}(\bm{\theta}) &= \DKL(p(\bm{x})\|p_{\bm{\theta}}(\bm{x}))
= - \Esp_{\bm{x}\sim p(\bm{x})}[-\log p_{\bm{\theta}}(\bm{x})] + \mathrm{const.} \nn \\
&= - \Esp_{\bm{x}\sim p(\bm{x})}[\log \pi(T_{\bm{\theta}}^{-1}(\bm{x}))+ \log |\det J_{T_{\bm{\theta}}^{-1}}(\bm{x})|] + \mathrm{const.} 
\end{align}
In practice, computing the expectation $\Esp_{\bm{x}\sim p(\bm{x})}$ is not possible 
and is replaced by a Monte Carlo technique using the data samples $\{\bm{x}^{(i)}\}_{i<N}$ (dropping the constant)
\begin{equation}
\mathcal{L}(\bm{\theta}) = -\frac{1}{N}\sum_{i=0}^{N-1} \left\{ \log \pi(T_{\bm{\theta}}^{-1}(\bm{x}^{(i)})) + \log |\det J_{T_{\bm{\theta}}^{-1}}(\bm{x}^{(i)})| \right\} 
\end{equation}
The computation of the  determinant of the Jacobian and its gradient may be challenging but more straightforwardly if the Jacobian matrices $\{J_{T_i}^{-1}\}_{i<n}$ (Eq.~\ref{eq-flow-jacob}) are triangular. Notice that we need $T^{-1}$ as well as the Jacobian and the gradients to minimize $\mathcal{L}(\bm{\theta})$ and obtain the  best $\theta$, but we need also $T_\theta$ for the generation of new $\bm{x}$ samples from sampling $z\sim \pi(z)$.

Different architectures of normalizing flows allowing such program are discussed in the review by \cite{Papamakarios2021}, among them the \textit{auto-regressive flows} are used for instance in the \texttt{Glow} image generator.

%
\subsection{Differences between the models}
%
%
\section{Experiment}
